---
title: "lab_03"
author: "Derek Willis/ Rob Wells"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse

library(tidyverse)
#install.packages('tidyverse')
# Load janitor
#install.packages('janitor')
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008) |> 
  clean_names()
# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv") |> clean_names()

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. 

Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1:
I was able to find four records while doing my search.

```{r}
pg_schools <- payments |>
  filter(
    str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") |
    str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOLS") |
      str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL") |
    str_detect(vendor_name, "PRINCE GEORGES PUBLIC SCHOOLS")) 

```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2:
The 2020 fiscal year has the largest total amount with $1362372202 in payments. I find it interesting that the year that the COVID-19 pandemic began has the highest payments of the fiscal years shown.
```{r}
pg_schools |> 
  group_by(`fiscal_year`) |> 
  summarize(count = n(), total = sum(amount, na.rm=TRUE)) |> 
  arrange(`fiscal_year`)
```

Q3. This question is in two parts. 

Part 1: Code: You're interested in examining the vendors paid for capital projects at the University of Maryland. Write code that does the following: 
--Filters payments data to capital projects at the University of Maryland
--Produce a list of vendors and the total amount they received. 

Part 2: Provide a written answer, not code.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? 
--Provide at least three examples of vendor names that are similar and should be combined into one. 
--Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?


A3.
1. I would add the zip code search in the code and set it to where it would be one business per zip code. 
2. Aerosol Monitoring & analysis Inc. and Aersol Monitoring and Analysis Inc. should be grouped together. All of the A&I Inc. vendors should be grouped together instead of separate. This also goes for A Morton Thomas & Associates Inc. 
3. I would group all the vendors with the same name (as some of them are similar but spelled differently) and combine all of their totals into one. 
```{r}
cap_umd <- payments |>
 filter(agency_name == 'CAPITAL PROJECTS- UNIVERSITY OF MARYLAND')

cap_umd |> 
  group_by(vendor_name) |>
  summarise(total = sum(amount, na.rm=TRUE))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. 

Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. 
Then create a total amount of money for each fiscal year. 

Write a sentence summarizing how stem cell research funding has changed over time. Be specific with the totals.

A4:
Over the years there was been a lot of fluctuation in how much funding does into stem cell research. 2021 had the least amount of funding with $4832225, and oddly enough the earliest year available, 2009, has the highest funding with $14643721 going towards stem cell research. It seems to have a trend between the years 2009-2020. It would fluctuate up and down anually which I find very interesting. 
```{r}
md_grants_loans |>
mutate(description = str_to_upper(description)) |>
  filter(str_detect(description, "STEM CELL RESEARCH ")) |>
    group_by(fiscal_year) |>
  summarise(total = sum(amount, na.rm=TRUE))
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?
20742
A5:
Under the University of Maryland Zip Code, there are 31 grantees. To clean up the results I would add a function to add up all of the ammounts together to give me a singular summed up number. the most intersting to me is the commercialization investment from OVALTZ INC. and Haystack Solutions Inc. Everything else is going to some kind of foundation or research while these are to promote.
```{r}
md_grants_loans |>
 mutate(zip5 = str_sub(zip_code, start=1L, end=5L)) |>
  mutate(zip5 = case_when(
    zip5 == "Vario" ~ NA,
    zip5 == "UB7 O" ~ NA,
    zip5 == "UB7 " ~ NA,
    .default = zip5
  ))
```
