---
title: "lab_01"
author: "Derek Willis / Rob Wells"
date: "2025-09-10"
output: html_document
---

PUT YOUR NAME AND DATE ON THIS ASSIGNMENT
Sophia Hernandez-Piña September 10, 2025


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? **A1**
The Computer, Math, and Natural Sciences department has the highest mean number of seats for its classes which is 440.00. I am honestly not surprised about the top departments shown because engineering, technology and the natural sciences are booming in the modern industry as technology is advancing, so inturn you would want more students in the departments to boost the number of graduates in the field. 
```{r}
umd_courses |>
  group_by(department) %>% 
summarize(mean_seats = mean(seats)) %>% 
arrange(desc(mean_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your result should sort to show the greatest number of classes first. If several classes are tied for the most offered, describe how many classes are ranked in the top category for offerings.
**A2**

```{r}
umd_courses %>% 
  group_by(title) %>% 
  summarize(
    count_classes = n()
  ) %>% 
  arrange(desc(count_classes))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most? HINT: To filter out blank and missing values from a column, drop_na() is your friend.
**A3**
Keith Yearwood taught the most classes.
```{r}
umd_courses %>% 
  drop_na(instructors) %>% 
  group_by(department,instructors) %>% 
  summarize(
    count_classes = n(),
    total_seats = sum(seats),
    max_seats = max(seats)
  ) %>% 
  arrange(desc(count_classes))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4**
I think that using drop_na was extremely helpful in accomplishing this question. It narrowed down the instructors column so I did not have to scroll through n/a input. I am fairly confident in my answer, as drap_na simplified the way to track down what instructors had how many classes. 
**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5**
Question: which instructor offered the most classes?
Answer: there are two instructors that are TBA at the top with 97 and 83 classes. The first named professor with the most classes offered is Keith Yearwood with 79 classes. 
```{r}
umd_courses %>% 
  drop_na(instructors) %>% 
  group_by(department,instructors) %>% 
  summarize(
    count_classes = n(),
    total_seats = sum(seats),
    max_seats = max(seats)
  ) %>% 
  arrange(desc(count_classes))
```
