---
title: "Reverse Engineering Project"
author: "Raphael Romero Ruiz & Sophia Hernadez-Piña"
date: "`Oct. 9 2025`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

13-23: Here we installed a package called tidycensus with enabled us to more easily use and work with data from the U.S. Census Bureau. We used this to pull information about the murder rates v. how many murders have been solved during certain time frames. We also installed the census_api_key, which holds the original data we will be using.
```{r}
install.packages("tidycensus")
```
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)
library(tidycensus)
library(readxl)

census_api_key("a053b1e13b17cfecd7fea96af2fc15059d046f8c", install=TRUE)
```
27-28: This code reads a CSV file that hold homicide data and then simplified the column names into a consistent, lowercase format, which made it easier for us to work with the data. 
```{r}
homicides_wapo <- read_csv("data-homicides/homicide-data.csv") %>% 
  clean_names()
```
33-39: This code fixes two issues in the reported dates and then converted all the data from the numbers into a proper date formation that R can understand. The mutate() function corrects two data entry errors, changing "201511105" to "20151105" and "201511018" to "20151118," while leaving all other dates the same, then we used the ymd() function to transform the numbers into a year-month-day format into actual date object that can be used for time-based analysis.
```{r}
# First make sure reported_date is in proper date format
homicides_wapo <- homicides_wapo %>%
  mutate(reported_date = case_when(
    reported_date == 201511105 ~ 20151105,
    reported_date == 201511018 ~ 20151118, 
    TRUE ~ reported_date
  )) %>%
  mutate(reported_date = ymd(reported_date))
```

47-60: This code imports historical Census population data for Baltimore city from a file, cleaning it up and reshaping it into a simple two-column format that shows the year and the population. The code standardizes the column names, filters to keep only the information from Baltimore city, and then extracts data from 2007, 2008 and 2009 populations. This made it easier to combine with other datasets.

```{r}
# Get population estimates for Baltimore City (2007-2014)
# First get 2007-2009 from XLS file
census_pop_2007_2009 <- read_xls("census-data/co-est00int-01-24.xls") %>%
  slice((6):(n() - 8)) %>%
  clean_names() %>%
  filter(str_detect(table_with_row_headers_in_column_a_and_column_headers_in_rows_3_through_4_leading_dots_indicate_sub_parts, "Baltimore city")) %>%
  transmute(
    pop_2007 = as.numeric(x10),
    pop_2008 = as.numeric(x11), 
    pop_2009 = as.numeric(x12)
  ) %>%
  pivot_longer(cols = everything(),
               names_to = "year_col", 
               values_to = "population") %>%
  mutate(year = as.numeric(str_extract(year_col, "[0-9]{4}"))) %>%
  select(year, population)
```
65-92: This uses the tidycensus package to download population estimates for Baltimore city from 2010 through 2014 directly from the U.S. Census' API. The get_estimates() function retrives the time-based population data using Baltimore's FIPS code (510), the the code translates the Census Bureau's numeric date codes into actual years, keeping only the estimates from 2010 to 2014, and creates a two-column dataset with year and population that matches the format of the earlier data.
```{r}
# Then get 2010-2014 from Census API
census_pop_2010_2014 <- get_estimates(
    geography = "county",
    vintage = 2019,
    time_series = TRUE,
    state = "MD",
    county = "510",  # Baltimore City FIPS code
    variable = "POP"
  ) %>%
  # The date column uses Census codes based on official documentation
  mutate(year = case_when(
    DATE == 1 ~ 2010,  # April 1, 2010 Census count
    DATE == 2 ~ NA,  # April 1, 2010 estimates base (not used)
    DATE == 3 ~ NA,  # July 1, 2010 estimate (not used)
    DATE == 4 ~ 2011,  # July 1, 2011 estimate
    DATE == 5 ~ 2012,  # July 1, 2012 estimate
    DATE == 6 ~ 2013,  # July 1, 2013 estimate
    DATE == 7 ~ 2014,  # July 1, 2014 estimate
    DATE == 8 ~ 2015,  # July 1, 2015 estimate
    DATE == 9 ~ 2016,  # July 1, 2016 estimate
    DATE == 10 ~ 2017, # July 1, 2017 estimate
    DATE == 11 ~ 2018, # July 1, 2018 estimate
    DATE == 12 ~ 2019, # July 1, 2019 estimate
    TRUE ~ as.numeric(DATE)
  )) %>%
  select(year, value) %>%
  rename(population = value) %>%
  filter(year >= 2010 & year <= 2014) %>%
  arrange(year)
```
97-101: This combines the two serparate population datasets, one covering 2007 to 2009 and the other covering 2010 to 2014, into a single dataset that sorts it chronologically by year. The bind_rows() function stacks the two datasets on top of each other snce they share the same column structure. arrange() makes sure all the years appear in order from earliest to latest, creating one continuous population series for Baltimore city from 2007 to 2014.
```{r}
# Combine 2007-2009 Excel data with 2010-2014 Census API data
pop_estimates_pre_2015 <- bind_rows(
  census_pop_2007_2009,
  census_pop_2010_2014
) %>%
  arrange(year)
```

106-135:This downloads population estimates for Baltimore city from 2015 to 2017 using the same Census API. It coverts the data into a clean format in chronological order. The process is similar to the previous download but filters the data to only keep the years 2015 through 2017, translating the Census' numeric date codes into readable years and organizes the data into two-column datasets. The dataset shows the year and population that can be joined with the pre-2015 data.

```{r}
# Get population estimates for Baltimore City (post-2015)
pop_estimates_post_2015 <- get_estimates(
    geography = "county",
    vintage = 2019,
    time_series = TRUE,
    state = "MD",
    county = "510",  # Baltimore City FIPS code
    variable = "POP"
  ) %>%
  # The date column uses Census codes based on official documentation
  mutate(year = case_when(
    DATE == 1 ~ 2010,  # April 1, 2010 Census count
    DATE == 2 ~ 2010,  # April 1, 2010 estimates base
    DATE == 3 ~ 2010,  # July 1, 2010 estimate
    DATE == 4 ~ 2011,  # July 1, 2011 estimate
    DATE == 5 ~ 2012,  # July 1, 2012 estimate
    DATE == 6 ~ 2013,  # July 1, 2013 estimate
    DATE == 7 ~ 2014,  # July 1, 2014 estimate
    DATE == 8 ~ 2015,  # July 1, 2015 estimate
    DATE == 9 ~ 2016,  # July 1, 2016 estimate
    DATE == 10 ~ 2017, # July 1, 2017 estimate
    DATE == 11 ~ 2018, # July 1, 2018 estimate
    DATE == 12 ~ 2019, # July 1, 2019 estimate
    TRUE ~ as.numeric(DATE)
  )) %>%
  select(year, value) %>%
  rename(population = value) %>%
  filter(year >= 2015 & year <= 2017) %>%
  arrange(year)
```

142-151: This analyzes baltimore homicide data before 2015, calculating the total number of homicides, arrests and the arrest rate for each year. The code filters the Washington Post homicide dataset to include only Baltimore cases before Jan 1, 2015, extracting the year from each homicide's reported date and then groups the data by year to count the total homicides. It then tallies how many cases resulted in arrests and calculate the percentage of homicides that led to arrests, rounding the arrest rate and presenting the data in chronological order.

```{r}
# Get Baltimore homicides and arrests by year before 2015
num_hom_and_arrest_bmore_by_year_pre_2015 <- homicides_wapo %>%
  filter(city == "Baltimore" & reported_date < "2015-01-01") %>%
  mutate(year = year(reported_date)) %>%
  group_by(year) %>%
  summarise(
    homicides = n(),
    arrests = sum(disposition == "Closed by arrest"),
    arrest_rate = round((sum(disposition == "Closed by arrest") / n()) * 100, 1)
  ) %>%
  arrange(year)  # Sort by year
```

158-167: This shows the same homicide and arrest analysis as the previous code but focuses specifically on Baltimore cases from 2015 through 2017, the period after Freddie Gray's death and the turmoil that followed. The code filters the Washington Post data to include only Baltimore homicides that occured between jan. 1, 2015 and Dec. 31, 2017. It then calculates the annual totals for homicides and arrests along with the arrest rate for each year, letting us compare how policing patterns may have changed during this timeframe. 

```{r}
# Get Baltimore homicides and arrests for 2015-2017
num_hom_and_arrests_bmore_by_year_post_2015 <- homicides_wapo %>%
  filter(city == "Baltimore" & reported_date >= "2015-01-01" & reported_date <= "2017-12-31") %>%
  mutate(year = year(reported_date)) %>%
  group_by(year) %>%
  summarise(
    homicides = n(),
    arrests = sum(disposition == "Closed by arrest"),
    arrest_rate = round((sum(disposition == "Closed by arrest") / n()) * 100, 1)
  ) %>%
  arrange(year)
```

174-176: This code combines the pre-2015 Baltimore homicide statisitics with population data to calculate the homicide rater ber 100,000 residents for each year. The lef_join() function merges the homicde and arrest counts with the corresponding population estimates by matching years. Then, the code calculates the standardized homicide rate by dividing the number of homicides by the populatio and multiplying by 100,000, rounding the results.

```{r}
# Calculate pre-2015 rates
bmore_hom_rates_pre_2015 <- num_hom_and_arrest_bmore_by_year_pre_2015 %>%
  left_join(pop_estimates_pre_2015, by = "year") %>%
  mutate(homicide_rate_per_100k = round((homicides / population) * 100000, 1))
```

183-185: This merges the 2015-2017 Baltimore homicide statistics with population data to calculate the homicide rate per 100,000 resident for those post-Freddie Gray years. Using the same methods for the pre-2015 analysis, the code joins the homicide and arrest counts with the corresponding population estimates by year. It then standardizes the homicide rate, divinding homicides by population and multiplying by 100,000, to create a rate that accounts for population changes that can be directly compared with the earlier years to asses whether violence in baltimore city has increased after 2015. 

```{r}
# Calculate post-2015 rates
bmore_hom_rates_post_2015 <- num_hom_and_arrests_bmore_by_year_post_2015 %>%
  left_join(pop_estimates_post_2015, by = "year") %>%
  mutate(homicide_rate_per_100k = round((homicides / population) * 100000, 1))
```

192-196: This combines the pre-2015 and post-2015 Baltimore homicide rate datasets into one complete dateset going from 2007 to 2017 and puts it in chronological order The bind_rows() function stacks the two datasets into a single dataset while arrange() ensures all the years are ordered sequentially, creating one continous time frame that allows for easy analysis of how Baltimore's homicide rates and arrest rates changed before and after 2015.

```{r}
# Combine all years into one comparison
bmore_rates_all_years <- bind_rows(
  bmore_hom_rates_pre_2015,
  bmore_hom_rates_post_2015
) %>%
  arrange(year)
```

```{r}
#everything below here is for assignment 3 and beyond
```

208-235: This downloads annual population estimates for Chicago from 2015 through 2017 using the Census API, giving the demographic data needed to calculate homicide rates in comparison with Baltimore. the get_estimates() function retrives Chicago's population using its place code (1400), filtering it to ensure its capturing the correct city rather than suburbs. It then converts the Census Bureau's data into readable years, then it keeps only the 2015-2017 estimates and organizes them into two-columns with year and population that can be merged with Chicago homicide data for a parallel analysis to Baltimore's post-2015 trends. 

```{r}
# CHICAGO ANALYSIS
# Get population estimates for Chicago (2015-2017)
pop_estimates_chicago_post_2015 <- get_estimates(
    geography = "place",
    vintage = 2019,
    time_series = TRUE,
    state = "IL",
    place = 1400,
    variable = "POP"
  ) %>%
  filter(str_detect(NAME, "^Chicago city, Illinois$")) %>%
  mutate(year = case_when(
    DATE == 1 ~ 2010,
    DATE == 2 ~ 2010,
    DATE == 3 ~ 2010,
    DATE == 4 ~ 2011,
    DATE == 5 ~ 2012,
    DATE == 6 ~ 2013,
    DATE == 7 ~ 2014,
    DATE == 8 ~ 2015,
    DATE == 9 ~ 2016,
    DATE == 10 ~ 2017,
    DATE == 11 ~ 2018,
    DATE == 12 ~ 2019,
    TRUE ~ as.numeric(DATE)
  )) %>%
  select(year, value) %>%
  rename(population = value) %>%
  filter(year >= 2015 & year <= 2017) %>%
  arrange(year)
```

242-251: This analyzes Chicago homicide data from 2015 to 2017, calculating the total number of homicides, arrests and the arrest rate for each year to create a comparison dataset with Baltimore. This code filters the Washington Post homicide data to include only Chicago cases between Jan. 1, 2015 and Dec. 31, 2017, and extracts the year from each homicide's reported date. It then groups the data by year to count total homicides, tally cases closed by arrest and calculate the percentage of homicides that resulted in arrests. This provides a parallel analysis that allows us to see whether baltimore's post-2015 trends were unique or showed a larger national pattern in urban policing. 

```{r}
# Get Chicago homicides for 2015-2017
num_hom_and_arrests_chicago_by_year_post_2015 <- homicides_wapo %>%
  filter(city == "Chicago" & reported_date >= "2015-01-01" & reported_date <= "2017-12-31") %>%
  mutate(year = year(reported_date)) %>%
  group_by(year) %>%
  summarise(
    homicides = n(),
    arrests = sum(disposition == "Closed by arrest"),
    arrest_rate = round((sum(disposition == "Closed by arrest") / n()) * 100, 1)
  ) %>%
  arrange(year)
```

258-260: This merges Chicago's 2015-2017 homicide statistics with its population data to calculate the city's homicide rate per 100,000 residents for those years. The left_join() function combines the homicide and arrest counts with the corresponding population estimates by years then calculates the homicide rate by dividng the number of homicides by the population and multiplying by 100,000, creating a metric that can be directly compared with Baltimore's rates to determine whether both cities had similar increases in violence during the same period of time.  

```{r}
# Calculate post-2015 rates
chi_hom_rates_post_2015 <- num_hom_and_arrests_chicago_by_year_post_2015 %>%
  left_join(pop_estimates_chicago_post_2015, by = "year") %>%
  mutate(homicide_rate_per_100k = round((homicides / population) * 100000, 1))
```

268-277: This analuzes Boston homicide data before 2015, calculating the total number of homicides, arrests and arrest rate for each year. The code filters the Washington Post homicide dataset to have only Boston cases before Jan 1, 2015, and extracts the year from each homicide's reported date. It then groups the data by year to count the total amount of homicides, tallies the cases closed by arrests and calculates the percent of homicides that resulted in arrest.

283-310: This downloads annual population estimates for Boston from 2015 to 2017 using the Census API, giving the demographic data needed to calculate homicicde rates for another comparison city. The get_estimates() fnction gets Boston's population using its place code (07000) in Massachusetts, converting the Census' date codes into readable years, and then keeps only the 2015-2017 estimates. It then organizes them into a two-column format with year and population that can be put together with Boston homicide data, allowing us to compare whether Baltimore's post-2015 homicide rate changes were different or similar to trends in Chicago and Boston.

```{r}
# BOSTON ANALYSIS
# Get Boston homicides and arrests by year (pre-2015)
num_hom_and_arrest_boston_by_year_pre_2015 <- homicides_wapo %>%
  filter(city == "Boston" & reported_date < "2015-01-01") %>%
  mutate(year = year(reported_date)) %>%
  group_by(year) %>%
  summarise(
    homicides = n(),
    arrests = sum(disposition == "Closed by arrest"),
    arrest_rate = round((sum(disposition == "Closed by arrest") / n()) * 100, 1)
  ) %>%
  arrange(year)
```
```{r}
# Get population estimates for Boston (2015-2017)
pop_estimates_boston_post_2015 <- get_estimates(
    geography = "place",
    vintage = 2019,
    time_series = TRUE,
    state = "MA",
    place = "07000",  # Boston place code
    variable = "POP"
  ) %>%
  mutate(year = case_when(
    DATE == 1 ~ 2010,
    DATE == 2 ~ 2010,
    DATE == 3 ~ 2010,
    DATE == 4 ~ 2011,
    DATE == 5 ~ 2012,
    DATE == 6 ~ 2013,
    DATE == 7 ~ 2014,
    DATE == 8 ~ 2015,
    DATE == 9 ~ 2016,
    DATE == 10 ~ 2017,
    DATE == 11 ~ 2018,
    DATE == 12 ~ 2019,
    TRUE ~ as.numeric(DATE)
  )) %>%
  #filter(str_detect(NAME, "^Boston, Massachusetts$")) %>% 
  select(year, value) %>%
  rename(population = value) %>%
  filter(year >= 2015 & year <= 2017) %>%
  arrange(year)
```

318-328: This analyzes Boston homicide data from 2015 to 2017, calculating the total number of homicides, arrests and arrest rate for each year to complete the the comparison dataset with Baltimore and Chicago. The code filters the Washington Post homicide data to include only Boston cases between Jan. 1m 2015, and Dec. 31, 2017. It extracts the year from each homicide;s date, and then groups the data by year to count the total homicides, tallies cases closed by arrest and calculates the percentage of homicides that resulted in arrests. This let us assess whether the patterns in Baltimore during this period were unique to the city or showed a broader trend across multiple police departments. 

332-334: This merges Boston's 2015-2017 homicide statistics with its population data to calculate the city's homicide rate per 100,000 residents for those years. The left_join() function combines the homicide and arrest counts with the population estimates by matching years. Then, it calculates the standardized homicide rate by dividing the number of homicicides by the population and multiplying by 100,000, creating a rate that accounts for Boston's population size and can be compared with Baltimore's and Chicago's rates to determine if the spike in violence after 2015 was a Baltimore-specific event or part of a wider trend affecting other cities. 

```{r}
# Get Boston homicides (2015-2017)
num_hom_and_arrests_boston_by_year_post_2015 <- homicides_wapo %>%
  filter(city == "Boston" & reported_date >= "2015-01-01" & reported_date <= "2017-12-31") %>%
  mutate(year = year(reported_date)) %>%
  group_by(year) %>%
  summarise(
    homicides = n(),
    arrests = sum(disposition == "Closed by arrest"),
    arrest_rate = round((sum(disposition == "Closed by arrest") / n()) * 100, 1)
  ) %>%
  arrange(year)
```
```{r}
# Calculate post-2015 rates
bos_hom_rates_post_2015 <- num_hom_and_arrests_boston_by_year_post_2015 %>%
  left_join(pop_estimates_boston_post_2015, by = "year") %>%
  mutate(homicide_rate_per_100k = round((homicides / population) * 100000, 1))
```

342-351: This analyzes St. Louis homicide data before 2015, calculating the total number of homicides, arrests and the arrest rate for each year to establish a baseline for another comparison city. The code filters the Washington Post homicide dataset to include only St. Louis cases before Jan. 1, 2015. It then takes the year from each homicide's reported date and then groups the data by year to count total homicides, tally cases closed by arrest and calculate the precentage of homicides that resulted in arrests. 

```{r}
# ST. LOUIS ANALYSIS
# Get St. Louis homicides and arrests by year before 2015
num_hom_and_arrest_stl_by_year_pre_2015 <- homicides_wapo %>%
  filter(city == "St. Louis" & reported_date < "2015-01-01") %>%
  mutate(year = year(reported_date)) %>%
  group_by(year) %>%
  summarise(
    homicides = n(),
    arrests = sum(disposition == "Closed by arrest"),
    arrest_rate = round((sum(disposition == "Closed by arrest") / n()) * 100, 1)
  ) %>%
  arrange(year)
```

358-385: This downloaded the annual population estimates for St. Louis from 2015 to 2017 using the Census API, giving the demographics needed to calculate homicide rates for this city. The get_estimates() function retrieves St. Louis' population using its place code (65000) in Missouri, converting the Census' date codes into readable years. It only keeps the 2015-2017 estimates and organizes them into a two-column format with year and population that can be merged with St. Louis homicide data. 

```{r}
# Get population estimates for St. Louis
pop_estimates_stl_post_2015 <- get_estimates(
    geography = "place",
    vintage = 2019,
    time_series = TRUE,
    state = "MO",
    place = "65000",  # St. Louis place code
    variable = "POP"
  ) %>%
  mutate(year = case_when(
    DATE == 1 ~ 2010,
    DATE == 2 ~ 2010,
    DATE == 3 ~ 2010,
    DATE == 4 ~ 2011,
    DATE == 5 ~ 2012,
    DATE == 6 ~ 2013,
    DATE == 7 ~ 2014,
    DATE == 8 ~ 2015,
    DATE == 9 ~ 2016,
    DATE == 10 ~ 2017,
    DATE == 11 ~ 2018,
    DATE == 12 ~ 2019,
    TRUE ~ as.numeric(DATE)
  )) %>%
  #filter(str_detect(NAME, "^St. Louis, Missouri$")) %>%
  select(year, value) %>%
  rename(population = value) %>%
  filter(year >= 2015 & year <= 2017) %>%
  arrange(year)
```

394-403:This analyzes St. Louis homicide data from 2015 through 2017, calculating the total number of homicides, arrests and arrest rate for each year to complete the post-2015 comparison with Baltimore, Chicago and Boston. The code filters the Washington Post homicide dataset to have only St. Louis cases between Jan. 1, 2015, and Dec. 31, 2017. The code then extracts the year from each homicide's reported date, and groups the data by year to count total homicides, tally cases closed by arrest and calculate the percentage of homicides that resulted in arrest. 

407-409: This combines St. Louis/ 2015-2017 homicide statistics with its population data to calculate the city's homicide rate per 100,000 residents for those years. The left_join() function combines the homicide and arrest counts with the corresponding population estimates by matching years, and then calculating the homicide rate by using the same formula for Baltimore, Chicago and Boston's information. 

```{r}
# Get St. Louis homicides for 2015-2017
num_hom_and_arrests_stl_by_year_post_2015 <- homicides_wapo %>%
  filter(city == "St. Louis" & reported_date >= "2015-01-01" & reported_date <= "2017-12-31") %>%
  mutate(year = year(reported_date)) %>%
  group_by(year) %>%
  summarise(
    homicides = n(),
    arrests = sum(disposition == "Closed by arrest"),
    arrest_rate = round((sum(disposition == "Closed by arrest") / n()) * 100, 1)
  ) %>%
  arrange(year)
```
```{r}
# Calculate post-2015 rates
stl_hom_rates_post_2015 <- num_hom_and_arrests_stl_by_year_post_2015 %>%
  left_join(pop_estimates_stl_post_2015, by = "year") %>%
  mutate(homicide_rate_per_100k = round((homicides / population) * 100000, 1))
```

416-422: this combines the 2015-2017 homicide rate dat from all four cities, Baltimore, Chicago, Boston and St. Louis, into a single dataset for comparative analysis. The bind_rows() function stacks the four separate datasets on top of each other after adding a label for each city using mutate(). Then we used arrange() to sort the combined data first by city name in alphabetical orger and then by year per city, creating one master dataset that allowed us to compare homicide rates, arrest rates and population trends accros all of these cities post-2015.In Chicago, the homicide arrest rate dropped 21 percentage points compared to 2014, in Boston 12 points, and in St. Louis it dropped 9 points. 

```{r}
# Combine all cities' post-2015 data for comparison
all_cities_post_2015 <- bind_rows(
  chi_hom_rates_post_2015 %>% mutate(city = "Chicago"),
  bos_hom_rates_post_2015 %>% mutate(city = "Boston"),
  stl_hom_rates_post_2015 %>% mutate(city = "St. Louis"),
  bmore_hom_rates_post_2015 %>% mutate(city = "Baltimore")
) %>%
  arrange(city, year)
```
